Ship-based lidar evaluation of Southern Ocean clouds in the storm-resolving general circulation model ICON, and the ERA5 and MERRA-2 reanalyses

### 2024-12-11 manuscript_agu.tex
Addressing comments

Increasing climate model resolution is one way of improving model accuracy of representation of the climate system \citeA{mauritsen2022}. It has been practiced since the advent of climate modeling as more computational power, memory, and storage capacity become available. It is, however, often not as easy as changing the grid size because of the complex interplay between model dynamics and physics, which necessitates adjusting and tuning all components together. Increasing resolution, is of course, limited by the available computational power and a trade-off with increasing parameterization complexity, which is another way of improving model accuracy. Current computational availability and acceleration from general-purpose computing on graphics processing units (GPUs) is progressing to enable km-scale (also called k-scale) Earth system models (ESMs) and coupled atmosphere--ocean general circulation models (AOGCMs) in research conditions today and operationally in the forthcoming years. Therefore, it represents a natural advance in climate modeling. Global storm-resolving models (GSRMs) are emerging as a new front in the development of high-resolution global climate models, with horizontal grid resolutions of about 2--8 km \citeA{satoh2019,stevens2019}. This is enough to resolve mesoscale convective storms, but smaller-scale convective plumes and cloud structure remain unresolved. At an approximately 5-km scale, non-hydrostatic processes also become important \citeA{weisman1997}, and for this reason such models are generally non-hydrostatic. The terms global cloud-resolving models or global convection-permitting/-resolving models are also sometimes used interchangeably with GSRMs but imply that clouds or convection are resolved explicitly, which is not entirely true for GSRMs, as this would require an even higher horizontal resolution \citeA{satoh2019}. Representative of these efforts is the DYnamics of the Atmospheric general circulation Modeled On Non-hydrostatic Domains (DYAMOND) project \citeA{stevens2019,dyamond}, which is an intercomparison of nine global GSRMs over two 40-day time periods in summer (1 August -- 10 September 2016) and winter (20 January -- 1 March 2020). A new one-year GSRM intercomparison is currently proposed by \citeA{takasuka2024}, with the hope of also evaluating the seasonal cycle and large-scale circulation. An alternative to using a computationally costly GSRM is to train an artificial neural network on GSRM output and use it for subgrid-scale clouds, as done with the GSRM ICON by \citeA{grundner2022} and \citeA{grundner2023}.

nextGEMS is a project \citeA{nextgems} focused on the research and development of GSRMs at multiple modeling centers and universities in Europe. The project also develops GSRM versions of the Icosahedral Nonhydrostatic Weather and Climate Model (ICON; \citeA{hohenegger2023}), the Integrated Forecasting System [IFS; \citeA{ifs2023}], and their ocean components at eddy-resolving resolutions: ICON-O \citeA{korn2022} coupled with ICON and Finite-Element/volumE Sea ice-Ocean Model [FESOM; \citeA{wang2014a}] and Nucleus for European modeling of the Ocean [NEMO; \citeA{madec2023}] coupled with IFS. The project has so far produced ICON and IFS simulations with three development versions called Cycle 1--3 and a pre-final version, with a final production version planned by the end of the project. nextGEMS is not the only project developing GSRMs; other GSRMs (or GSRM versions of climate models) currently in development include: Convection-Permitting Simulations With the E3SM Global Atmosphere Model [SCREAM; \citeA{caldwell2021}], Atmospheric Model [NICAM; \citeA{satoh2008}], Unified Model (UM), eXperimental System for High-resolution modeling for Earth-to-Local Domain [X-SHiELD; \citeA{shield}], Action de Recherche Petite Echelle Grande Echelle-NonHydrostatic version [ARPEGE-NH; \citeA{bubnova1995,voldoire2017}], Finite-Volume Dynamical Core on the Cubed Sphere [FV3, \citeA{lin2004}], the National Aeronautics and Space Administration (NASA) Goddard Earth Observing System global atmospheric model version 5 [GEOS5; \citeA{putman2011}], Model for Prediction Across Scales [MPAS; \citeA{skamarock2012}], and System for Atmospheric Modeling [SAM; \citeA{khairoutdinov2003}].

Multiple cloud properties have an effect on shortwave (SW) and longwave (LW) radiation. To first order, the total cloud fraction, cloud phase, and the liquid and ice water path are the most important cloud properties influencing SW and LW radiation. These properties are in turn influenced by the atmospheric thermodynamics, convection and circulation, and both the indirect and direct effects of aerosols. Second-order effects on SW and LW radiation are associated with the cloud droplet size distribution, ice crystal habit, cloud lifetime, and direct radiative interaction with aerosols. In the 6$^\mathrm{th}$ phase of the Coupled Model Intercomparison Project [CMIP6; \citeA{eyring2016}], the cloud feedback has increased relative to CMIP5 \citeA{zelinka2020}, which is one of the main reasons for the higher climate sensitivity of CMIP6 models.

The Southern Ocean (SO) is known to be a problematic region for climate model biases \citeA{schuddeboom2021,hyder2018,cesana2022,zhao2022} due to a lack of surface and in situ observations and being a lower priority region for numerical weather prediction (NWP) and climate model development because of its distance from populated areas. Nevertheless, radiation biases and changes over an area of its size have a substantial influence on the global climate \citeA{rintoul2011}, such as affecting the Earth radiation balance, ocean heat, and carbon uptake \citeA{williams2023}, and the SO is an important part of the global ocean conveyor belt \citeA{wang2014b}. In general, marine clouds have a disproportionate effect on top of atmosphere (TOA) SW radiation due to the relatively low albedo of the sea surface. The relative longitudinal symmetry of the SO means that model cloud biases tend to be similar across longitudes.

@@

SO radiation biases have been relatively large and systematic compared to the rest of the globe since at least CMIP3 \citeA{trenberth2010}, and the SO SW cloud radiative effect (CRE) bias is still positive in eight analyzed CMIP6 models analyzed by \citeA{schuddeboom2021} over the high-latitude SO, whereas over the low-latitude SO it tends to be more neutral or negative in some models. Too much absorbed SW radiation over the SO was also identified in the GSRM SCREAM \citeA{caldwell2021}. Compensating biases are possible, such as the ``too few too bright'' cloud bias, characterized by too small cloud fraction and too large cloud albedo \citeA{wall2017,kuma2020}, previously described by \citeA{webb2001}, \citeA{weare2004}, \citeA{zhang2005}, \citeA{karlsson2008}, \citeA{nam2012}, \citeA{klein2013}, and \citeA{bender2017} in other regions and models, which means that a model can maintain a reasonable SW radiation balance by reflecting too much SW radiation from clouds, but these cover too small an area. A study by \citeA{konsta2022} showed that this type of bias is still present in six analyzed CMIP6 models in tropical marine clouds, using the GCM-Oriented Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO) Cloud Product [CALIPSO--GOCCP; \citeA{chepfer2010}] and Polarization \& Anisotropy of Reflectances for Atmospheric Sciences coupled with Observations from a Lidar [PARASOL; \citeA{lier2008}] as a reference. They suggest improper simulation of subgrid-scale cloud heterogeneity as a cause. Compensating cloud biases in the Australian Community Climate and Earth System Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were analyzed by \citeA{fiddes2022} and \citeA{fiddes2024}. \citeA{possner2022} showed that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction on the order of 30\% [relative to Moderate Resolution Imaging Spectroradiometer (MODIS) data] and overestimates downwelling TOA SW radiation by approximately 10 Wm$^\mathrm{-2}$ [relative to the Clouds and the Earth’s Radiant Energy System (CERES) data] in the highest model resolution run (2.5 km). \citeA{zhao2022} reported a similar SW radiation bias in five analyzed CMIP6 models over high-latitude SO and the total cloud fraction underestimation on the order of 10\% over the entire 40--60°S SO. Recently, \citeA{ramadoss2024} analyzed 48 hours of km-scale ICON limited-area model NWP simulations over a SO region adjacent to Tasmania against the Clouds, Aerosols, Precipitation, Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN) voyage cloud and precipitation observations \citeA{mcfarquhar2021}. They found the ICON cloud optical thickness was underestimated relative to Himawari‐8 satellite observations but also identified large differences in cloud top phase.

In general, sea surface temperature (SST) biases in the SO can originate either in the atmosphere \citeA{hyder2018}, caused by too much shortwave heating of the surface, too little longwave cooling of the surface, or in the ocean circulation. Interactions of both are also possible, for example, SST affecting clouds and clouds affecting the surface radiation. Using the European Centre for Medium-Range Weather Forecasts (ECMWF) Reanalysis 5 (ERA5) as a reference, \citeA{zhang2023} have shown that SST biases have improved in CMIP6 compared to CMIP5, with SST overall increasing in the later CMIP phase. However, over the SO this resulted in an even higher positive bias, especially in the Atlantic Ocean (AO) sector of the SO, increasing by up to 1°C. \citeA{luo2023} identified that the SO SST bias in an ensemble of 18 CMIP6 models originates not from the surface heat and radiation fluxes (using reanalyses as a reference), but from a warm bias in the Northern Atlantic Deep Water.

The main aim of this study is to evaluate the GSRM version of ICON. ICON is developed and maintained jointly by Deutscher Wetterdienst, Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Previous studies have identified substantial large-scale biases in climate model clouds over the SO, affecting sea surface temperature and the Earth’s albedo. Our aim is to quantify how well the GSRM ICON is simulating clouds in this region, particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \citeA{mace2009,medeiros2010} and ground clutter \citeA{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on CALIPSO (neither of which are now operational) are affected by the abovementioned issues, resulting in a strong underestimation of cloud occurrence below 2 km relative to ground-based lidar observations \citeA{mcerlich2021}. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parametrized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the RV \emph{Polarstern} and other voyages and stations as a reference for the model evaluation.

Altogether, we analyzed about 2400 days of data from 31 voyages and one sub-Antarctic station covering diverse longitudes and latitudes of the SO. To achieve a like-for-like comparison with the model, we used a ground-based lidar simulator called the Automatic Lidar and Ceilometer Framework [ALCF; \citeA{kuma2021}]. We contrasted the results with ERA5 \citeA{era5} and the Modern-Era Retrospective analysis for Research and Applications, Version 2 [MERRA-2; \citeA{gelaro2017}].

@@

Together, we analyzed data from 31 voyages of RV \emph{Polarstern}, the resupply vessel (RSV) \emph{Aurora Australis}, RV \emph{Tangaroa}, RV \emph{Nathaniel B. Palmer}, Her (now His) Majesty's New Zealand Ship (HMNZS) \emph{Wellington} and one sub-Antarctic station (Macquarie Island) in the SO south of 40°S between 2010 and 2021. Fig. \ref{fig:map} shows a map of the voyages, Table \ref{tab:voyages} lists the voyages, campaigns, and stations, and Table \ref{tab:voyage-references} lists references where available. Altogether, the voyages and station dataset comprised 2421 days of data south of 40°S, but the availability of ceilometer data was slightly shorter due to gaps in measurements.

Missing days in the ceilometer data were HMNZSW16 (7 days): 24--27 November, 10 December, 16--17 December 2016; Measurements of Aerosols, Radiation, and CloUds over the Southern Ocean (MARCUS; 3 days): 8, 10 November, 10 December 2017; Macquarie Island Cloud Radiation Experiment (MICRE; 9 days): 7--8, 29 June, 5, 16 July, 15 August, 17 October 2016, 11 February, 21 March 2017; TAN1502 (1 day): 24 January.

The data sources contained ceilometer observations captured by the Vaisala CL51 and CT25K, and the Lufft CHM 15k, described in detail below (Sections \ref{sec:cl51} and \ref{sec:chm15k}). A ceilometer is a low-power, near-infrared vertically pointing lidar principally designed to measure cloud base, but they also measure the full vertical structure of clouds as long as the laser signal is not attenuated by thick clouds, which can be used to infer additional information such as a cloud mask and cloud occurrence by height. We note that during the MICRE campaign, the ceilometers Vaisala CT25K and CL51 were installed at the Macquarie Island station concurrently, but in our analysis we only used the CT25K data obtained from the Atmospheric Radiation Measurement (ARM) data archive.

Apart from lidar observations, radiosondes were launched on weather balloons at regular synoptic times on the RV \emph{Polarstern}, MARCUS, NBP17024, TAN1702, and TAN1802 voyages and campaigns, measuring pressure, temperature, relative humidity, and the global navigation satellite system coordinates. Derived thermodynamic (virtual potential temperature, lifting condensation level, etc.) and dynamic physical quantities (wind speed and direction) for the measured vertical profiles were calculated with rstool \citeA{rstool}. Surface meteorological quantities were measured continuously by an onboard automatic weather station or individual instruments.

>>>>

Increasing climate model resolution is one way of improving model accuracy of representation of the climate system \cite{mauritsen2022}. It has been practiced since the advent of climate modeling as more computational power, memory, and storage capacity become available. It is, however, often not as easy as changing the grid size because of the complex interplay between model dynamics and physics, which necessitates adjusting and tuning all components together. Increasing resolution, is of course, limited by the available computational power and a trade-off with increasing parameterization complexity, which is another way of improving model accuracy. Current computational availability and acceleration from general-purpose computing on graphics processing units (GPUs) is progressing to enable km-scale (also called k-scale) Earth system models (ESMs) and coupled atmosphere--ocean general circulation models (AOGCMs) in research conditions today and operationally in the forthcoming years. Therefore, it represents a natural advance in climate modeling. Global storm-resolving models (GSRMs) are emerging as a new front in the development of high-resolution global climate models, with horizontal grid resolutions of about 2--8 km \cite{satoh2019,stevens2019}. This is enough to resolve mesoscale convective storms, but smaller-scale convective plumes and cloud structure remain unresolved. At an approximately 5-km scale, non-hydrostatic processes also become important \cite{weisman1997}, and for this reason such models are generally non-hydrostatic. The terms global cloud-resolving models or global convection-permitting/-resolving models are also sometimes used interchangeably with GSRMs but imply that clouds or convection are resolved explicitly, which is not entirely true for GSRMs, as this would require an even higher horizontal resolution \cite{satoh2019}. Representative of these efforts is the DYnamics of the Atmospheric general circulation Modeled On Non-hydrostatic Domains (DYAMOND) project \cite{stevens2019,dyamond}, which is an intercomparison of nine global GSRMs over two 40-day time periods in summer (1 August -- 10 September 2016) and winter (20 January -- 1 March 2020). A new one-year GSRM intercomparison is currently proposed by \citeA{takasuka2024}, with the hope of also evaluating the seasonal cycle and large-scale circulation. An alternative to using a computationally costly GSRM is to train an artificial neural network on GSRM output and use it for subgrid-scale clouds, as done with the GSRM ICON by \citeA{grundner2022} and \citeA{grundner2023}.

nextGEMS is a project \cite{nextgems} focused on the research and development of GSRMs at multiple modeling centers and universities in Europe. The project also develops GSRM versions of the Icosahedral Nonhydrostatic Weather and Climate Model (ICON; \citeA{hohenegger2023}), the Integrated Forecasting System [IFS; \citeA{ifs2023}], and their ocean components at eddy-resolving resolutions: ICON-O \cite{korn2022} coupled with ICON and Finite-Element/volumE Sea ice-Ocean Model [FESOM; \citeA{wang2014a}] and Nucleus for European modeling of the Ocean [NEMO; \citeA{madec2023}] coupled with IFS. The project has so far produced ICON and IFS simulations with three development versions called Cycle 1--3 and a pre-final version, with a final production version planned by the end of the project. nextGEMS is not the only project developing GSRMs; other GSRMs (or GSRM versions of climate models) currently in development include: Convection-Permitting Simulations With the E3SM Global Atmosphere Model [SCREAM; \citeA{caldwell2021}], Atmospheric Model [NICAM; \citeA{satoh2008}], Unified Model (UM), eXperimental System for High-resolution modeling for Earth-to-Local Domain [X-SHiELD; \citeA{shield}], Action de Recherche Petite Echelle Grande Echelle-NonHydrostatic version [ARPEGE-NH; \citeA{bubnova1995,voldoire2017}], Finite-Volume Dynamical Core on the Cubed Sphere [FV3, \citeA{lin2004}], the National Aeronautics and Space Administration (NASA) Goddard Earth Observing System global atmospheric model version 5 [GEOS5; \citeA{putman2011}], Model for Prediction Across Scales [MPAS; \citeA{skamarock2012}], and System for Atmospheric Modeling [SAM; \citeA{khairoutdinov2003}].

Multiple cloud properties have an effect on shortwave (SW) and longwave (LW) radiation. To first order, the total cloud fraction, cloud phase, and the liquid and ice water path are the most important cloud properties influencing SW and LW radiation. These properties are in turn influenced by the atmospheric thermodynamics, convection and circulation, and both the indirect and direct effects of aerosols. Second-order effects on SW and LW radiation are associated with the cloud droplet size distribution, ice crystal habit, cloud lifetime, and direct radiative interaction with aerosols. In the 6$^\mathrm{th}$ phase of the Coupled Model Intercomparison Project [CMIP6; \citeA{eyring2016}], the cloud feedback has increased relative to CMIP5 \cite{zelinka2020}, which is one of the main reasons for the higher climate sensitivity of CMIP6 models.

The Southern Ocean (SO) is known to be a problematic region for climate model biases \cite{schuddeboom2021,hyder2018,cesana2022,zhao2022} due to a lack of surface and in situ observations and being a lower priority region for numerical weather prediction (NWP) and climate model development because of its distance from populated areas. Nevertheless, radiation biases and changes over an area of its size have a substantial influence on the global climate \cite{rintoul2011}, such as affecting the Earth radiation balance, ocean heat, and carbon uptake \cite{williams2023}, and the SO is an important part of the global ocean conveyor belt \cite{wang2014b}. In general, marine clouds have a disproportionate effect on top of atmosphere (TOA) SW radiation due to the relatively low albedo of the sea surface. The relative longitudinal symmetry of the SO means that model cloud biases tend to be similar across longitudes.

@@

SO radiation biases have been relatively large and systematic compared to the rest of the globe since at least CMIP3 \cite{trenberth2010}, and the SO SW cloud radiative effect (CRE) bias is still positive in eight analyzed CMIP6 models analyzed by \citeA{schuddeboom2021} over the high-latitude SO, whereas over the low-latitude SO it tends to be more neutral or negative in some models. Too much absorbed SW radiation over the SO was also identified in the GSRM SCREAM \cite{caldwell2021}. Compensating biases are possible, such as the ``too few too bright'' cloud bias, characterized by too small cloud fraction and too large cloud albedo \cite{wall2017,kuma2020}, previously described by \citeA{webb2001}, \citeA{weare2004}, \citeA{zhang2005}, \citeA{karlsson2008}, \citeA{nam2012}, \citeA{klein2013}, and \citeA{bender2017} in other regions and models, which means that a model can maintain a reasonable SW radiation balance by reflecting too much SW radiation from clouds, but these cover too small an area. A study by \citeA{konsta2022} showed that this type of bias is still present in six analyzed CMIP6 models in tropical marine clouds, using the GCM-Oriented Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO) Cloud Product [CALIPSO--GOCCP; \citeA{chepfer2010}] and Polarization \& Anisotropy of Reflectances for Atmospheric Sciences coupled with Observations from a Lidar [PARASOL; \citeA{lier2008}] as a reference. They suggest improper simulation of subgrid-scale cloud heterogeneity as a cause. Compensating cloud biases in the Australian Community Climate and Earth System Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were analyzed by \citeA{fiddes2022} and \citeA{fiddes2024}. \citeA{possner2022} showed that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction on the order of 30\% and overestimates net dowward TOA SW radiation by approximately 10 Wm$^\mathrm{-2}$ in the highest model resolution run (2.5 km). \citeA{zhao2022} reported a similar SW radiation bias in five analyzed CMIP6 models over high-latitude SO and the total cloud fraction underestimation on the order of 10\% over the entire 40--60°S SO. Recently, \citeA{ramadoss2024} analyzed 48 hours of km-scale ICON limited-area model NWP simulations over a SO region adjacent to Tasmania against the Clouds, Aerosols, Precipitation, Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN) voyage cloud and precipitation observations \cite{mcfarquhar2021}. They found the ICON cloud optical thickness was underestimated relative to Himawari‐8 satellite observations but also identified large differences in cloud top phase.

In general, sea surface temperature (SST) biases in the SO can originate either in the atmosphere \cite{hyder2018}, caused by too much shortwave heating of the surface, too little longwave cooling of the surface, or in the ocean circulation. Interactions of both are also possible, for example, SST affecting clouds and clouds affecting the surface radiation. Using the European Centre for Medium-Range Weather Forecasts (ECMWF) Reanalysis 5 (ERA5) as a reference, \citeA{zhang2023} have shown that SST biases have improved in CMIP6 compared to CMIP5, with SST overall increasing in the later CMIP phase. However, over the SO this resulted in an even higher positive bias, especially in the Atlantic Ocean (AO) sector of the SO, increasing by up to 1°C. \citeA{luo2023} identified that the SO SST bias in an ensemble of 18 CMIP6 models originates not from the surface heat and radiation fluxes (using reanalyses as a reference), but from a warm bias in the Northern Atlantic Deep Water.

The main aim of this study is to evaluate the GSRM version of ICON. ICON is developed and maintained jointly by Deutscher Wetterdienst, Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Previous studies have identified substantial large-scale biases in climate model clouds over the SO, affecting sea surface temperature and the Earth’s albedo. Our aim is to quantify how well the GSRM ICON is simulating clouds in this region, particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on CALIPSO (neither of which are now operational) are affected by the abovementioned issues, resulting in a strong underestimation of cloud occurrence below 2 km relative to ground-based lidar observations \cite{mcerlich2021}. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parametrized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the RV \emph{Polarstern} and other voyages and stations as a reference for the model evaluation.

Altogether, we analyzed about 2400 days of data from 31 voyages and one sub-Antarctic station covering diverse longitudes and latitudes of the SO. To achieve a like-for-like comparison with the model, we used a ground-based lidar simulator called the Automatic Lidar and Ceilometer Framework [ALCF; \citeA{kuma2021}]. We contrasted the results with ERA5 \cite{era5} and the Modern-Era Retrospective analysis for Research and Applications, Version 2 [MERRA-2; \citeA{gelaro2017}].

@@

Together, we analyzed data from 31 voyages of RV \emph{Polarstern}, the resupply vessel (RSV) \emph{Aurora Australis}, RV \emph{Tangaroa}, RV \emph{Nathaniel B. Palmer}, Her (now His) Majesty's New Zealand Ship (HMNZS) \emph{Wellington} and one sub-Antarctic station (Macquarie Island) in the SO south of 40°S between 2010 and 2021. Fig. \ref{fig:map} shows a map of the campaigns, Table \ref{tab:voyages} lists the campaigns, and Table \ref{tab:voyage-references} lists references where available. Altogether, the analyzed dataset comprised 2421 days of data south of 40°S, but the availability of ceilometer data was slightly shorter due to gaps in measurements.

The campaigns contained ceilometer observations captured by the Vaisala CL51 and CT25K, and the Lufft CHM 15k, described in detail below (Sections \ref{sec:cl51} and \ref{sec:chm15k}). A ceilometer is a low-power, near-infrared vertically pointing lidar principally designed to measure cloud base, but they also measure the full vertical structure of clouds as long as the laser signal is not attenuated by thick clouds, which can be used to infer additional information such as a cloud mask and cloud occurrence by height. We note that during the MICRE campaign, the ceilometers Vaisala CT25K and CL51 were installed at the Macquarie Island station concurrently, but in our analysis we only used the CT25K data obtained from the Atmospheric Radiation Measurement (ARM) data archive.

Apart from lidar observations, radiosondes were launched on weather balloons at regular synoptic times on the RV \emph{Polarstern}, MARCUS, NBP17024, TAN1702, and TAN1802 campaigns, measuring pressure, temperature, relative humidity, and the global navigation satellite system coordinates. Derived thermodynamic (virtual potential temperature, lifting condensation level, etc.) and dynamic physical quantities (wind speed and direction) for the measured vertical profiles were calculated with rstool \cite{rstool}. Surface meteorological quantities were measured continuously by an onboard automatic weather station or individual instruments.

\begin{table}[p!]
\caption{
An overview of the analyzed campaigns (voyages and stations). Start, end, and the number of days (UTC; inclusive) refer to the time period when the vessel was south of 40°S. Abbreviations: ceilometer (ceil.), Australia (AU), New Zealand (NZ), South America (SA), Atlantic Ocean (AO), and Africa (AF). The number of days is rounded to the nearest integer. CL51/31 indicates CL51 configured to emulate CL31. Missing days in the ceilometer data were HMNZSW16 (7 days): 24--27 November, 10 December, 16--17 December 2016; MARCUS (3 days): 8, 10 November, 10 December 2017; MICRE (9 days): 7--8, 29 June, 5, 16 July, 15 August, 17 October 2016, 11 February, 21 March 2017; TAN1502 (1 day): 24 January.
}
\label{tab:voyages}
\centering
\small
\begin{tabular}{llllllr}
\textbf{Name} & \textbf{Vessel or station} & \textbf{Ceil.} & \textbf{Region} & \textbf{Start} & \textbf{End} & \textbf{Days}\\
\hline
AA15-16  & RSV \emph{Aurora Australis}   & CL51    & AU       & 2015-10-22 & 2016-02-22 & 124 \\
HMNZSW16 & HMNZS \emph{Wellington}       & CHM 15k & NZ       & 2016-11-23 & 2016-12-19 & 27 \\
MARCUS   & RSV \emph{Aurora Australis}   & CT25K   & AU       & 2017-10-29 & 2018-03-26 & 149 \\
MICRE    & Macquarie Is. station         & CT25K   & AU/NZ    & 2016-04-03 & 2018-03-14 & 710 \\
NBP1704  & RV \emph{Nathaniel B. Palmer} & CHM 15k & NZ       & 2017-04-14 & 2017-06-08 & 55 \\
PS77/2   & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2010-12-01 & 2011-02-04 & 65 \\
PS77/3   & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2011-02-07 & 2011-04-14 & 66 \\
PS79/2   & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2011-12-06 & 2012-01-02 & 27 \\
PS79/3   & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2012-01-10 & 2012-03-10 & 61 \\
PS79/4   & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2012-03-14 & 2012-04-08 & 26 \\
PS81/2   & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2012-12-02 & 2013-01-18 & 47 \\
PS81/3   & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2013-01-22 & 2013-03-17 & 55 \\
PS81/4   & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2013-03-18 & 2013-04-16 & 30 \\
PS81/5   & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2013-04-20 & 2013-05-23 & 33 \\
PS81/6   & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2013-06-10 & 2013-08-12 & 63 \\
PS81/7   & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2013-08-15 & 2013-10-14 & 60 \\
PS81/8   & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2013-11-12 & 2013-12-14 & 31 \\
PS81/9   & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2013-12-21 & 2014-03-02 & 71 \\
PS89     & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2014-12-05 & 2015-01-30 & 56 \\
PS96     & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2015-12-08 & 2016-02-14 & 68 \\
PS97     & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2016-02-15 & 2016-04-06 & 52 \\
PS103    & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2016-12-18 & 2017-02-02 & 46 \\
PS104    & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2017-02-08 & 2017-03-18 & 39 \\
PS111    & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2018-01-21 & 2018-03-14 & 52 \\
PS112    & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2018-03-18 & 2018-05-05 & 49 \\
PS117    & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2018-12-18 & 2019-02-07 & 51 \\
PS118    & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2019-02-18 & 2019-04-08 & 50 \\
PS123    & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2021-01-10 & 2021-01-31 & 21 \\
PS124    & RV \emph{Polarstern}          & CL51    & SA/AO/AF & 2021-02-03 & 2021-03-30 & 55 \\
TAN1502  & RV \emph{Tangaroa}            & CL51/31 & NZ       & 2015-01-20 & 2015-03-12 & 51 \\
TAN1702  & RV \emph{Tangaroa}            & CHM 15k & NZ       & 2017-03-09 & 2017-03-31 & 23 \\
TAN1802  & RV \emph{Tangaroa}            & CHM 15k & NZ       & 2018-02-07 & 2018-03-20 & 41 \\
\hline
\textbf{Total} &                         &         &          &            &            & \textbf{2421}\\
\hline
\end{tabular}
\normalsize
\end{table}

\begin{table}[t!]
\caption{Campaign publication references.}
\label{tab:voyage-references}
\centering
\small
\begin{tabular}{lp{14.5cm}}
\textbf{Name} & \textbf{References}\\
\hline
AA15-16  & \citeA{klekociuk2020} \\
MARCUS   & \citeA{mcfarquhar2021,xia2024,niu2024} \\
MICRE    & \citeA{mcfarquhar2021} \\
NBP1704  & \citeA{ackley2020} \\
PS77/2   & \citeA{kniglanglo2011a,kniglanglo2011b,kniglanglo2011c,kniglanglo2014a,fahrbach2011} \\
PS77/3   & \citeA{kniglanglo2011d,kniglanglo2011e,kniglanglo2012a,kniglanglo2014b,knust2011} \\
PS79/2   & \citeA{kniglanglo2012b,kniglanglo2012c,kniglanglo2012d,kniglanglo2014c,kattner2012} \\
PS79/3   & \citeA{kniglanglo2012e,kniglanglo2012f,kniglanglo2012g,kniglanglo2014d,wolfgladrow2012} \\
PS79/4   & \citeA{kniglanglo2012h,kniglanglo2012i,kniglanglo2012j,kniglanglo2014e,lucassen2012} \\
PS81/2   & \citeA{kniglanglo2013a,kniglanglo2013b,kniglanglo2013c,kniglanglo2014f,boebel2013} \\
PS81/3   & \citeA{kniglanglo2013d,kniglanglo2013e,kniglanglo2013f,kniglanglo2014g,gutt2013} \\
PS81/4   & \citeA{kniglanglo2013g,kniglanglo2013h,kniglanglo2013i,kniglanglo2014q,bohrmann2013} \\
PS81/5   & \citeA{kniglanglo2013j,kniglanglo2013k,kniglanglo2013l,kniglanglo2014r,jokat2013} \\
PS81/6   & \citeA{kniglanglo2013m,kniglanglo2013n,kniglanglo2013o,kniglanglo2014h,lemke2013} \\
PS81/7   & \citeA{kniglanglo2013p,kniglanglo2013q,kniglanglo2014i,kniglanglo2016a,meyer2013} \\
PS81/8   & \citeA{kniglanglo2013r,kniglanglo2014j,kniglanglo2014k,kniglanglo2014l,schlindwein2014} \\
PS81/9   & \citeA{kniglanglo2014m,kniglanglo2014n,kniglanglo2014o,kniglanglo2014p,knust2014} \\
PS89     & \citeA{kniglanglo2015a,kniglanglo2015b,kniglanglo2015c,kniglanglo2015d,boebel2016}\\
PS96     & \citeA{kniglanglo2016b,kniglanglo2016c,kniglanglo2016d,kniglanglo2016e,schrder2017} \\
PS97     & \citeA{kniglanglo2016f,kniglanglo2016g,kniglanglo2016h,kniglanglo2016i,lamy2017} \\
PS103    & \citeA{kniglanglo2017a,kniglanglo2017b,kniglanglo2017c,kniglanglo2017d,boebel2018} \\
PS104    & \citeA{kniglanglo2017e,kniglanglo2017f,kniglanglo2017g,gohl2018,schmithsen2021a} \\
PS111    & \citeA{schmithsen2019a,schmithsen2020a,schmithsen2021b,schmithsen2021c,schrder2018} \\
PS112    & \citeA{schmithsen2019b,schmithsen2020b,schmithsen2021d,schmithsen2021e,meyer2018} \\
PS117    & \citeA{schmithsen2019c,schmithsen2020c,schmithsen2021f,schmithsen2021g,boebel2019} \\
PS118    & \citeA{schmithsen2019d,schmithsen2020d,schmithsen2021h,schmithsen2021i,dorschel2019} \\
PS123    & \citeA{schmithsen2021j,schmithsen2021m,schmithsen2021n,schmithsen2021k,hoppmann2023a} \\
PS124    & \citeA{schmithsen2021o,schmithsen2021q,schmithsen2021p,hoppmann2023b} \\
TAN1802  & \citeA{kremser2020,kremser2021} \\
\hline
\end{tabular}
\end{table}
